---
title: "Launching RStudion in Docker"
output:
  dcTemplate::dc_lesson_template:
    fig_width: 6
    fig_height: 6
    highlight: pygments
---


```{r knitr_init, echo = FALSE, cache = FALSE}
library(knitr)

## Global options
options(max.print = "75")
opts_chunk$set(cache = TRUE,
               prompt = FALSE,
               tidy = TRUE,
               comment = "> #",
               message = FALSE,
               warning = FALSE)
opts_knit$set(width = 75)
```

## Lesson Objectives

- Launch RStudio inside of a Docker container
- Linking a volume to a Docker container
- Load into the container plotting it

## Installation

First things first: **install Docker**: on [Linux](https://docs.docker.com/linux/step_one/), [Mac](https://docs.docker.com/mac/step_one/), or [Windows](https://docs.docker.com/windows/step_one/). These install guides link to a bunch of introductory material after installation is complete; it's not necessary to complete those tutorials for this lesson, but they are an excellent introduction to basic Docker usage.

## Launching RStudio in Docker

The first thing we need to do to launch Docker is to open a Unix Shell. If you're on Mac or Windows, in the last step you installed something called the *Docker Quickstart Terminal*; open that up now - it should look like just a plain shell prompt (`~$`), but really it's pointing at a linux virtual machine that Docker likes to run in, and this is where you should do everything for the rest of this tutorial unless otherwise noted. If you're on a linux machine, then you can just use a plain old terminal prompt.

Next, we will ask Docker to run an image that already exists, we will use the rstudio Docker image from [Rocker](https://github.com/rocker-org/rocker/wiki) which will allow us to run RStudio inside the container. 

**call-out box**
`p` and `--rm` are flags that allow you to customize how you run the container. `p` tells Docker that you will be using a port (the location of which we specify afterwards as `8787:8787`). This is necessary to run RStudio in the container. Finally, --rm ensures that when we quit the container, the container is deleted. If we did not do this, everytime we run a container, a version of it will be saved to our local computer. This can lead to the eventual wastage of a lot of disk space until we manually remove these containers. Later we will show you how to save your container if you want to do so.

~~~
docker run --rm -p 8787:8787 rocker/rstudio
~~~

**call-out box**
If you try to run a Docker container which you have not installed locally then Docker will automatically search for the container on Docker Hub (an online repository for docker images) and download it if it exists.

The command above will lead RStudio-Server to launch invisibly. To connect to it, open a browser and enter `http://`, followed by your ip address, followed by `:8787`. If you are running a Mac or Windows machine, you will find the ip address on the first line of text that appeared in your terminal when you launched the Docker Quickstart Terminal. For example, you should see:

~~~

                        ##         .
                  ## ## ##        ==
               ## ## ## ## ##    ===
           /"""""""""""""""""\___/ ===
      ~~~ {~~ ~~~~ ~~~ ~~~~ ~~~ ~ /  ===- ~~~
           \______ o           __/
             \    \         __/
              \____\_______/


docker is configured to use the default machine with IP 192.168.99.100
For help getting started, check out the docs at https://docs.docker.com
~~~

Thus, you would enter `http://192.168.99.100:8787` in your browser as the url.

If you are running a Linux machine, you can use `localhost` as the ip address. For example: `http://localhost:8787`

This should lead to you being greeted by the RStudio welcome screen. Log in using:

username: rstudio
password: rstudio

Now you should be able to work with RStudio in your browser in much the same way as you would on your desktop.

Try to now look at your files of your virtual computer (docker container). Go to
file -> open file. You will see that there are actually no files. The reason for
this is that this image came with no files.

Next, open a new R Script, e.g. by going to file -> New file -> R Script. 
Enter the following code in the script, run it and save it.
```{r, eval=FALSE}
x <- 1:5
y <- 6:10

pdf("firstplot.pdf")
plot(x, y)
dev.off()
```
If you look at your files again now, you will see two files (the plot and the 
script).

Now, given that we used the --rm flag when we launched the Docker container, anything we create on the machine will be gone. Let's verify this. First, close the browser tab where you have RStudio open, and then go to your terminal window from where you launched the Docker container and type Contol+C.

Now relaunch your a Docker container using the rstudio image as you did previously, e.g., docker run --rm -p 8787:8787 rocker/rstudio in the terminal and typing `http://192.168.99.100:8787` in your browser window and see if the rscript and plot you saved is still there.

## Linking a volume to a Docker container

Well that leads us to the question of, how can we save our work if the container is deleted when we exit the container? One solution is to link a volume (for example your local hard drive) to the container so that you can access the data there as well as being able to save things there.

## Loading into the container plotting it


## Summary


## Challenge Questions
